Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-12-02 Manuscript/manuscriptPS.tex
Update text

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids interact with signaling and other proteins via
electrostatic interactions and direct binding, and induce
membrane fusion and phase separation together with calcium ions.
Molecular details of these phenomena are not well understood
because accurate models to interpret the experimental data has not
been available. Here, we collect a set of experimental NMR data which
could be used together with molecular dynamics (MD) simulations
to interpret the lipid headgroup structures and details of ion binding
in pure and mixed PS and PS:PC lipid bilayers. Aiming to interpret the
data, we use the open collaboration method to go through the available MD simulation models for PS lipids.
However, none of the models reproduce the experimental data with sufficient
accuracy to interpet the structural details of lipid headgroups or
ion binding details in lipid bilayers containing PS lipids.
In contrast to PC lipids, the tested MD simulation models do not
correctly reproduce the qualitative response of PS lipid headgroups
to the bound ions or changes in the lipid composition. 
Our results pave the way for the model improvement to correctly
describe negatively charged membranes and their interactions with ions.
\end{abstract}

@@

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16}.
However, the recent comparisons of PC lipid headgroup and glycerol backbone
C-H bond order parameters calculated from different simulation models
revealed that improvements in the current force fields are needed
to correctly reproduce the headgroup structure and ion binding to
lipid bilayers \cite{botan15,catte16,ollila16}. The ion binding affinity
to POPC bilayer was then improved by implicitly including the electronic
polarizability using the electronic continuum correction \cite{melcr18}.
Here, we collect the set of experimentally measured lipid headgroup and
glycerol backbone C-H bond order parameters, which can be used to
evaluate the quality of headgroup structure and the ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared against
the collected experimental data. The results pave the way for the
development of MD simulation force fields that correctly describe
PS lipid headgroup structure and its interactions with ions.
Such models are expected to be useful in elucidating the biological
role of PS and other lipid headgroups because glycerol backbone and
lipid headgroups behave similarly in model membranes and in
bacteria \cite{gally81,scherer87,seelig90}.

%In NMRlipids I and II project we were looking for a MD model
%which would correctly reproduce headgroup and glycerol
%backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
%Here we extend the same goal for lipids with negatively charged PS headgroup.
%Chemical structure of PS headgroup together with other common biological
%lipids is shown in Fig. \ref{lipids}.
%\begin{figure}[]

%  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
%  \caption{\label{lipids}
%    Chemical structures and labels for the headgroup carbons.
%  }
%\end{figure}

%The interpretation of this data and some other results has been that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}

%Phospholipids containing various polar headgroups and acyl
%chains are essential building blocks of biological membranes.
%Atomistic level structural details of lipids and lipid-ion 
%interactions are considered highly important in several
%biological processes. The lipid structure and ion binding
%can be studied in detail with NMR spectroscopy. However,
%the structural interpretation of NMR data requires usage
%of models. The combination of classical molecular dynamics
%simulations with NMR data, especially with C-H bond
%order parameters, can potentially give atomistic resolution
%interpretation of structure and dynamics of molecules \cite{botan15,ollila16,ferreira16}.

%Our recent studies concluded that MD models are cabable to
%give structural interpretation for phosphatidylcholine 
%hydrophopic acyl chain region, while hydrophilic headgrop, glycerol 
%backbone and cation binding posed a major challenge for current
%force fields \cite{botan15,ollila16,catte16,ferreira16}. 
%These conclusions were reached by reviewing extensively 
%available experimental data from various sources and using
%NMRlipids Open Collaboration project to collect massive 
%amount of simulation data \cite{botan15,catte16}.

%Here we apply the same approach in search of MD simulation
%models which would reproduce glycerol backbone and 
%headgroup structures of lipids with PE, PG and PS headgroups.  
%In addition, we attempt to find a MD simulation model
%that would be able to correctly describe cation binding in
%bilayer containing negatively charged PG and PS lipids.

%Several MD simulation models have succesfully described 
%acyl chain structures and their qaulitative changes in different
%conditions for phosphatidylcholine lipids (for review see \cite{ollila16}).
%However, the current simulation models were not found to be accurate
%enough for full structural interpretation of phosphatidylcholine headgroup and
%glycerol backbone \cite{botan15}. Also Na$ +$  binding affinities
%were found to be significantly overestimated in several models and 
%none of the available models was able to Ca$2+$

\begin{table*}[!htb]
\caption{List of MD simulations of pure PS bilayers without additional salt.
   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
   and CKP refers to the version with more Gromos compatible version
   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
}\label{PSsystems}
\begin{tabular}{l c r r r r r c c}
  %\hline
  % some footnotes are not visible in typeset-MS (pdf)
 lipid/counter-ions & force field for lipids / ions & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$  & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
    POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
    POPS/K$^+$   & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
    POPS/Na$^+$  & Berger \cite{??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
    POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
    POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\
    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
%\end{sidewaystable*} 

@@

\caption{List of POPC:POPS mixture simulations with different amounts of added ions. 
  The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
  these correspond the concentrations reported in the experiments by Roux et al.~\cite{roux90}.

@@

Headgroup and glycerol backbone C-H bond order parameters of POPS
were determined from chemical-shift resolved dipolar splittings
measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}
and corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The following experimental setups were used.

@@

\emph{NMR numerical simulations}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

@@

The C-H bond order parameters were calculated directly
from the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and the membrane normal.
Angular brackets point to the average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid molecule in the system. The average and
the standard error of the mean were then calculated over different lipids.
The number density profiles were calculated using {\it gmx density} tool
from Gromacs sofware package \cite{gromacsMANUAL}.

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking in PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%The detailed structural differences between the headgroups is, however, not known.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured in Ref.~\cite{ferreira16}.
  }
\end{figure}

>>>>

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids interact with signaling and other proteins via
electrostatic interactions and direct binding, and induce
membrane fusion and phase separation together with calcium ions.
Molecular details of these phenomena are not well understood
because accurate models to interpret the experimental data has not
been available. Here, we collect a set of experimental NMR data which
can be used together with molecular dynamics (MD) simulations
to interpret the lipid headgroup structures and details of ion binding
in pure PS and mixed PS:PC lipid bilayers. We use the open collaboration
method to collect data from available MD simulation models of PS lipids.
None of the models reproduce the NMR data with experimental accuracy,
but the best models suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
all the tested force fields overestimate the cation binding affinity to lipid bilayers containing
PS lipids. In contrast to PC lipids, the response of PS headgroups
to the bound ions qualitatively differs from experiments in the tested MD
simulation models. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe negatively charged membranes and their interactions with ions.
The work is performed in the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18}.
However, the headgroup structures and ion binding affinity 
predicted by the MD simulation force fields for bilayers containing PS lipids are not
evaluated against experiments. Recent studies comparing the
C-H bond order parameters between simulations and experiments show that the
structure of PC lipid headgroup and glycerol backbone are not well
captured by most simulation models \cite{botan15} and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}. Based on this data,
the cation binding affinity to POPC bilayer was then improved by implicitly
including the electronic polarizability using the electronic continuum correction \cite{melcr18}.

\begin{table*}[htb]
%\begin{sidewaystable*}[!p]
\centering
\caption{The list of MD simulations of pure PS bilayers without additional salt.
  Simulation details are given in the supplementary information
    %   The lipid force fields named as in our previous work~\cite{botan15}.
}\label{PSsystems}
%begin{minipage}[t]{\textwidth}
\begin{tabular}{l c r r r r r c c}
  %\hline
  % some footnotes are not visible in typeset-MS (pdf)
 lipid/counter-ions & force field for lipids / ions & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$  & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
  \hline
    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
    \hline
    POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
    POPS/K$^+$   & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
    POPS/Na$^+$  & Berger \cite{??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
    POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
    POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\
    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
    \end{tabular}
%\end{minipage}
%\end{sidewaystable*} 
\end{table*}

Here, we collect the set of experimentally measured lipid headgroup and
glycerol backbone C-H bond order parameters, which can be used to
evaluate the quality of headgroup structure and the ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared with
the collected experimental data using the NMRlipids open collaboration
project (\url{www.nmrlipids.blogspot.fi}). The results pave the way
for the development of lipid force fields with realistic description of
the headgroup region of negatively charged lipids in physiological salt
conditions. Such models are expected to be useful in understanding
biological function of lipid headgroups and glycerol backbone because
they behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.

@@

\caption{The list of POPC:POPS mixtures simulated with different amounts of added ions. 
  The salt concentrations are calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
  This corresponds the concentration in buffer before solvating lipids, which are
  reported in the experiments by Roux et al.~\cite{roux90}.
  The simulation details are given in the supplementary information.

@@

Headgroup and glycerol backbone C-H bond order parameters of POPS
were determined from the chemical-shift resolved dipolar splittings
measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}.
The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The following experimental setups were used.

@@

\emph{NMR numerical simulations}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

@@

The C-H bond order parameters were calculated directly
from the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and the membrane normal.
Angular brackets point to the average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid molecule in the system. The average and
the standard error of the mean were then calculated over different lipids.
The python program ({\it scripts/calcOrderParameters.py}) utilizing
MDAnalysis library \cite{agrawal11,gowers16} is available in Ref. \citenum{MATCHgit}. 
The number density profiles were calculated using {\it gmx density} tool
from Gromacs sofware package \cite{gromacsMANUAL}.

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking in PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%The detailed structural differences between the headgroups is, however, not known.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured in Ref.~\cite{ferreira16}.
  }
\end{figure}

### 2018-12-02 Manuscript/manuscriptPSsuppl.tex
Update text

\subsection{GROMOS-CKP}
\todo{To be written by Piggot}

>>>>

\subsection{GROMOS-CKP}
\todo{To be written by Piggot}
  CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
   and CKP refers to the version with more Gromos compatible version
   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
